---
title: "Chapter 2: Graphs (one variable)"
author: "Paul Northrop"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Chapter 2: Graphs (one variable)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat1004.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "", prompt = TRUE, collapse = TRUE)
devtools::load_all() 
```

The main purpose of this vignette is to provide R code to produce graphs that involve more than one variable.  In particular we consider scatterplots, which feature in Section 2.6 of the [STAT1004 notes](https://moodle.ucl.ac.uk/pluginfile.php/3054127/mod_resource/content/22/STAT1004notes-2x1.pdf).
See also the [Chapter 2: Graphs (one variable)](stat1004-ch2b-graphs-vignette.html).

The R code used in this vignette are available [here](stat1004-ch2b-graphs-2-vignette.R).
The code underlying the `box_plot` function is available at  [GitHub](https://github.com/paulnorthrop/stat1004/tree/master/R/graphs.R).

## The Oxford Birth Times data 

These data are available in the data frame `ox_births`.
```{r, message = FALSE}
library(stat1004)
birth_times <- ox_births[, "time"]
day <- ox_births[, "day"]
```

To display these data we manipulate them into a matrix that is of the same format as Table 2.1 in the notes.  The number of birth times varies between days so we pad the matrix with R's missing values code `NA` in order that each column of the matrix has the same number of rows.

```{r, message = FALSE}
ox_mat <- matrix(NA, ncol = 7, nrow = 16)
for (i in 1:7) {
  day_i_times <- ox_births$time[which(ox_births$day == i)]
  ox_mat[1:length(day_i_times), i] <- sort(day_i_times)
  colnames(ox_mat) <- paste("day", 1:7, sep = "")
}  
ox_mat
```

We have a numeric continuous variable, `birth_times`, and a categorical variable, `day`.  The following code produces the graphs in the [STAT1004 lecture slides](https://moodle.ucl.ac.uk/mod/resource/view.php?id=92031) that contain a plot of birth_times for each day of the week. 

```{r, fig.show = 'hold', fig.width = 3.4, fig.height = 2.75}
par(mar = c(4, 4, 0.5, 1))
xlab <- "time (hours)"
x_labs <- c(min(birth_times), pretty(birth_times), max(birth_times))
# top left
box_plot(birth_times~day, col = 8, ylab = xlab, pch = 16, xlab = "day")
# top right
box_plot(birth_times~day, col = 8, horizontal = TRUE,  axes = FALSE, xlab = xlab, ylab = "day", pch = 16)
axis(1, at = x_labs, labels = x_labs)
axis(2, at = 1:7, labels = 1:7, lwd = 0, lty = 0)
# bottom left
box_plot(birth_times~day, axes = FALSE, ylab = xlab, pch = 16, lty = 1, range = 0, boxcol = "white", staplewex = 0, medlty = "blank", medpch = 16, xlab = "day")
axis(1, at = 1:7, labels = 1:7, lwd = 0, lty = 0)
axis(2, at = x_labs, labels = x_labs)
# bottom right
box_plot(birth_times~day, horizontal = TRUE, axes = FALSE, xlab = xlab, pch = 16, lty = 1, range = 0, boxcol = "white", staplewex = 0, medlty = "blank", medpch = 16)
axis(1, at = x_labs, labels = x_labs)
axis(2, at = 1:7, labels = 1:7, lwd = 0, lty = 0, las = 1)
```

## The 2000 US Presidential Election data


