---
title: "Chapter 2: Graphs"
author: "Paul Northrop"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Chapter 2: Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat1004.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
devtools::load_all() 
```

The main purpose of this vignette is to provide R code to produce graphs that feature in Sections 2.4 and 2.5 of the [STAT1004 notes](https://moodle.ucl.ac.uk/pluginfile.php/3054127/mod_resource/content/22/STAT1004notes-2x1.pdf).

The R code used in this vignette are available [here](stat1004-ch2b-graphs-vignette.R).

## The Oxford Birth Times data 

These data are available in the data frame `ox_births`.
```{r, message = FALSE}
library(stat1004)
birth_times <- ox_births[, "time"]
sort(birth_times)
```

## Empirical cumulative distribution function

1. **Can you work out from the output below what the function `ecdf` does? You could also use `?ecdf`**

```{r, fig.width = 6, fig.height = 4}
ox_ecdf <- ecdf(birth_times)
ox_ecdf(1)
ox_ecdf(2)
2/95
ox_ecdf(4.7)
23/95
ox_ecdf(19)
ox_ecdf(100)
```

The `plot` method for an object returned by `ecdf` produces a plot much like (the hollow circles in) Figure 2.3 in the STAT1004 notes.

```{r, fig.width = 6, fig.height = 4}
plot(ox_ecdf, main = "", xlab = "time (hours)", ylab = "cumulative relative frequency", pch = 1)
```

## Histogram

For an explanation of how to construct a histogram, and why it is constructed in this way, see the short [How to draw a histogram video](https://moodle.ucl.ac.uk/mod/url/view.php?id=2085771).  The following code produces the plots on the right side of Figure 2.4 in the notes.  Note that in the first plot the `probability = TRUE` (or `prob = TRUE` for short) argument is important: it ensures that the total area of the rectangles of the histogram is equal to 1.   In the second plot it isn't necessary to include `prob = TRUE` because if the bin widths are unequal then (sensibly) `hist` uses `prob = TRUE`.

```{r, fig.show='hold', fig.width = 6.5}
hist(birth_times, prob = TRUE, col = 8, main = "", xlab = "time (hours)")
br <- c(seq(from = 0, to = 12, by = 2), 20)
hist(birth_times, breaks = br, col = 8, main = "", xlab = "time (hours)")
```

The `hist` function returns an object (a *list* of several objects in fact) containing the information used to construct the plot.  If we assign this to an R object, `ox_tab` say, then we can look at this information and perhaps use it.  The `ls` function tells us the names of the objects in this list.  Use `?hist` to find out what these objects are.

```{r}
ox_tab <- hist(birth_times, plot = FALSE)
ls(ox_tab)
```

We use the information in `ox_tab` to reproduce the plot in Figure 2.2 of the notes.

```{r, fig.width = 6.5, fig.height = 4}
cum_rel_freq <- cumsum(c(0, ox_tab$counts)) / length(birth_times)
plot(ox_tab$breaks, cum_rel_freq, type = "b", pch = 16, ylab = "cumulative relative frequency", xlab = "time (hours)")
abline(h = 1, lty = 2)
```

The following code produces the `histogram-like` alternative plots in the lecture slides [STAT1004 lecture slides](https://moodle.ucl.ac.uk/mod/resource/view.php?id=92031).  As you can see we can do many things to change the appearance of plots. Use `?par` to see a list of the graphical parameters that we can use.

```{r, fig.width = 6.5, fig.height = 4}
# adjust plot margins, produce a 2 by 2 array of plots (fill row 1 then row 2)
par(mar = c(4, 4, 1, 1), mfrow = c(2, 2))
# no vertical axis
hist(birth_times, col = 8, prob = TRUE, axes = FALSE, xlab = "time (hours)", ylab = "", main = "")
axis(1, line = 0.5)
# no vertical axis plus rug of points
hist(birth_times, col = 8, prob = TRUE, axes = FALSE, xlab = "time (hours)", ylab = "",main = "")
axis(1, line = 0.5)
rug(birth_times, line = 0.5, ticksize = 0.05)
# non-shaded frequency polygon
n <- length(ox_tab$mids)
ox_tab$mids <- c(ox_tab$mids[1], ox_tab$mids, ox_tab$mids[n])
ox_tab$density <- c(0, ox_tab$density, 0)
plot(ox_tab$mids, ox_tab$density, xlab = "time (hours)", ylab = "density", type = "l", las = 1) 
axis(1, line = 0)
# shaded frequency polygon with no vertical axis
plot(ox_tab$mids, ox_tab$density, xlab="time (hours)", ylab = "", type = "l", bty= "l", axes = FALSE)
axis(1, line = -0.4)
polygon(ox_tab$mids, ox_tab$density, col = 8)
```

## Stem-and-leaf plot

The short [How to draw a histogram video](https://moodle.ucl.ac.uk/mod/url/view.php?id=2085771) also describes how to draw a stem-and-leaf plot.

```{r}
# The default plot
stem(birth_times)
# The plot that appears in the notes
stem(birth_times, scale = 2)
```

## Dotplot

It is surprisingly difficult to produce nice-looking dotplot using standard R functions.  One possibility is to use `stripchart`.  The code below also produces the plots Figure 2.5 in the notes.

```{r, fig.width = 6.5, fig.height = 4}
stripchart(birth_times, method = "stack", pch = 16, at = 0, offset = 2/3)
title(xlab = "time (hours)")
x <- round(birth_times)
stripchart(x, method = "stack", pch = 16, at = 0, axes = FALSE, offset = 2/3)
title(xlab = "time (hours)")
x_labs <- c(min(x), pretty(x), max(x))
axis(1, at = x_labs)
```

## Boxplot

