---
title: "Chapter 5: binomial, geometric and Poisson distributions"
author: "Paul Northrop"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Chapter 5: binomial, geometric and Poisson distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat1004.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "", prompt = TRUE, collapse = TRUE)
#devtools::load_all()
```

```{r, include = FALSE}
# Set a random number `seed'.  
set.seed(2601)
```

The purposes of this vignette are to direct you to R functions and relating to the **discrete** probability distributions condsidered in Chapter 5 of the STAT0002 notes and to provide code to do some of the things that appear in the lecture slides.  

First, we do this using the standard R functions (in the **stats** package).  Then we repeat some of the calculations using the the **distributions3** package, which provides a neat way to work with random variables in R.

We illustrate these functions using the Australian birth times data, which are available in the data frame `aussie_births`.


```{r}
library(stat1004)
head(aussie_births)
```

The **stats** package contains functions to evaluate the probability mass function (p.m.f.), cumulative distribution function (c.d.f.) and quantile function (which could also be called the inverse c.d.f.) of many discrete probability distributions, including the binomial, geometric and Poisson.  It also contains functions to simulate random variates from these distributions.  For a general description of these functions use `?Distributions`.  These functions are named in the following way, where `xxx` is the name (perhaps abbreviated) of the distribution in question: `dxxx` (p.m.f), `pxxx` (c.d.f), `qxxx` (quantile function) and `rxxx` (random variate generation).

## The binomial distribution

Following the STAT0002 notes let us suppose that (before we observe the data) the number $Y$ of boy babies is a random variable with a binomial distribution(44, $p$) distribution, where $p$ is the probability that a randomly chosen birth produces a boy.  We use the data to estimate $p$.

```{r}
boys <- sum(aussie_births[, "sex"] == "boy")
girls <- sum(aussie_births[, "sex"] == "girl")
p_hat <- boys / (boys + girls)
p_hat
```

The function `dbinom` evaluates the p.m.f. of a binomial distribution.  See `?Binomial` for information about this function and `pbinom`, `qbinom` and `rbinom`.

The following code produces a plot like the ones we looked at  lectures.

```{r, fig.align='center', fig.width=7}
# Plot the binomial(44, 1/2) p.m.f. 
n <- 44
plot(0:n, dbinom(0:n, size = n, prob = 1 / 2), type = "h", 
     ylab = "P(X = x)", xlab = "x")
```

```{r, fig.align='center', fig.width=7}
#barplot(dbinom(0:44, 44, p_hat), names.arg = 0:44, width = 0.5, xlim = c(0, 44))
#plot(0:44, dbinom(0:44, 44, p_hat), type = "h")
```

For the purposes of illustrating how the functions `pbinom`, `qbinom` and `rbinom` let us suppose that $p = 1/2$.  If we wanted to calculate the probabilities $P(Y < 10)$  

```{r}
pbinom
```

1. **?**

```{r, fig.align='center', fig.width=7}
pjn <- rbinom(100000, 44, 1 / 2)
barplot(table(pjn))
```

## The geometric distribution

## The Poisson distribution

## The distributions3 package

In September 2019 the **distributions3** package [@distributions3] was released.
A neat feature of this package is that it enables us to create an R object that corresponds to a particular random variable.  Then we are able easily to use some simple R functions to perform calculations with this random variable.  The way that this works is perhaps more intuitive than the way in which we use functions like `dbinom()` above.

```{r}
library(distributions3)
# Create on R object that is, effectively, a binomial(n, 1/2) random variable
Y <- Binomial(n, 1 / 2)
Y
```

```{r, fig.align='center', fig.width=7}
# Reproduce the plot.  
plot(0:n, pmf(Y, 0:n), type = "h", ylab = "P(X = x)", xlab = "x")
```

## References

<script type="text/x-mathjax-config">
   MathJax.Hub.Config({  "HTML-CSS": { minScaleAdjust: 125, availableFonts: [] }  });
</script>
