---
title: "Chapter 2: Graphs (more than one variable)"
author: "Paul Northrop"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Chapter 2: Graphs (more than one variable)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: stat1004.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "", prompt = TRUE, collapse = TRUE)
devtools::load_all() 
```

The main purpose of this vignette is to provide R code to produce graphs that involve more than one variable.  We consider two general situations: (i) plotting the values of a continuous variable for different values of a categorical variable (see [The Oxford Birth Times data](#the-oxford-birth-times-data)) and (ii) scatter plots of oner variable against another (see [The 2000 US Presidential Election data](#the-2000-us-presidential-election-data)).  The latter feature in Section 2.6 of the [STAT1004 notes](https://moodle.ucl.ac.uk/pluginfile.php/3054127/mod_resource/content/22/STAT1004notes-2x1.pdf).  See also the [Chapter 2: Graphs (one variable)](stat1004-ch2b-graphs-vignette.html).

The R code used in this vignette are available [here](stat1004-ch2b-graphs-2-vignette.R).  The code underlying the `box_plot` function is available at  [GitHub](https://github.com/paulnorthrop/stat1004/tree/master/R/graphs.R).

## The Oxford Birth Times data 

These data are available in the data frame `ox_births`.
```{r, message = FALSE}
library(stat1004)
birth_times <- ox_births[, "time"]
day <- ox_births[, "day"]
```

To display these data we manipulate them into a matrix that is of the same format as Table 2.1 in the notes.  The number of birth times varies between days so we pad the matrix with R's missing values code `NA` in order that each column of the matrix has the same number of rows.

```{r, message = FALSE}
ox_mat <- matrix(NA, ncol = 7, nrow = 16)
for (i in 1:7) {
  day_i_times <- ox_births$time[which(ox_births$day == i)]
  ox_mat[1:length(day_i_times), i] <- sort(day_i_times)
  colnames(ox_mat) <- paste("day", 1:7, sep = "")
}  
ox_mat
```

We have a numeric continuous variable, `birth_times`, and a categorical variable, `day`.  The following code produces the graphs in the [STAT1004 lecture slides](https://moodle.ucl.ac.uk/mod/resource/view.php?id=92031) that contain a plot of birth_times for each day of the week. 

```{r, fig.show = 'hold', fig.width = 3.4, fig.height = 2.75}
par(mar = c(4, 4, 0.5, 1))
xlab <- "time (hours)"
x_labs <- c(min(birth_times), pretty(birth_times), max(birth_times))
# top left
box_plot(birth_times~day, col = 8, ylab = xlab, pch = 16, xlab = "day")
# top right
box_plot(birth_times~day, col = 8, horizontal = TRUE,  axes = FALSE, xlab = xlab, ylab = "day", pch = 16)
axis(1, at = x_labs, labels = x_labs)
axis(2, at = 1:7, labels = 1:7, lwd = 0, lty = 0)
# bottom left
box_plot(birth_times~day, axes = FALSE, ylab = xlab, pch = 16, lty = 1, range = 0, boxcol = "white", staplewex = 0, medlty = "blank", medpch = 16, xlab = "day")
axis(1, at = 1:7, labels = 1:7, lwd = 0, lty = 0)
axis(2, at = x_labs, labels = x_labs)
# bottom right
box_plot(birth_times~day, horizontal = TRUE, axes = FALSE, xlab = xlab, pch = 16, lty = 1, range = 0, boxcol = "white", staplewex = 0, medlty = "blank", medpch = 16)
axis(1, at = x_labs, labels = x_labs)
axis(2, at = 1:7, labels = 1:7, lwd = 0, lty = 0, las = 1)
```

## The 2000 US Presidential Election in Florida

These data are available in the data frame `USelection`.  See `?USelection` for details.

```{r}
# County identifiers and location
head(USelection[, 1:4])
# County demographic variables
head(USelection[, 5:12])
# Numbers of votes for candidates
head(USelection[, 13:22])
```

For the moment we simply produce some scatter plots.  A separate vignette will be devoted to these data.

A plot to show the locations of the counties.

```{r, fig.width = 6, fig.height = 5}
plot(-USelection[, "lon"], USelection[, "lat"], xlab = "longitude (degrees north)", ylab = "latitude (degrees east)", pch = 16)
```

**1. Can you see the outline of the state of Florida?**

A plot of the percentage of the vote for Buchanan against population size.  
```{r, fig.width = 6, fig.height = 5}
pbuch <- USelection$buch/USelection$tvot
pch <- 1 + 3 * (USelection[, "co_names"] == "PalmBeach")
pch
plot(USelection$npop, pbuch, xlab = "population", ylab = "Buchanan % vote", pch = pch)
```

**2. Can you see how we have managed to identify Palm Beach County by using a different plotting character?**

Pairwise scatter plots of the demographic variables.

```{r, fig.width = 7, fig.height = 7}
pairs(USelection[, 5:12])
```
